---
title: "Plot_functions"
output: html_document
date: "2025-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plotting functions
### Function that that shows distribution of one species at all GBIF location starting from ARMS location

First plot that is created uses a species name, the distances from one ARMS location to all GBIF locations and the output graph is put in a directory.The graph shows the distribution in a percentage against the total amount of occurrences. The x-axis is the distance in km.
Goal of this graph: View a possible pattern of movement for one species.

```{r cars}
country.final <- function(species, distances, output_dir) {
  max_x <- max(long_sea$x) #Put the maximum distance in a variable to use for the x-axis limit
  plot <- ggplot(long_sea, aes(x = x, fill = location)) +
    geom_histogram(aes(y = after_stat(count / sum(count) * 100)), # makes sure the y-axis changes from frequency to percentage
                   binwidth = 50, #set the bar width
                   boundary = 0, #Control bin alignment
                   position = "stack") + #how overlapping data is positione 
    labs(title = paste0("Frequencies of Sea distances for ", species," from all ARMS locations" ),
         x = "Sea distance (km)", y = "Percentage of total occurrences (%)") + #Set the axis labels 
    theme_bw() +
    #scale_fill_brewer(palette = "Set1") +  # You can choose a different palette if you like
    theme(plot.title = element_text(hjust = 0.5, size = 13, face = "bold"), # set title font size, placement
          plot.margin = margin(0.3, 0.3, 0.4, 0.4, "cm"), #Set margins for plot
          axis.text = element_text(size = 10),           # Set font size for axis numbers
          axis.title = element_text(size = 12), # Set axis titles size
          legend.title = element_text(size = 13),   # Increase legend title size
          legend.text = element_text(size = 10),    # Increase legend text size
          legend.key.size = unit(1, "lines"), # Sets the size of the symbol next to the legend text
          axis.text.x = element_text(angle = 45, hjust = 1)) +  # Sets angle of x-axis text 
    scale_x_continuous(breaks = seq(0, max_x*1.1, by = 500), expand = c(0, 0))+ #sets the behaviour of x-axis
    scale_y_continuous(expand = c(0, 0)) + #sets behaviour for y axis
    coord_cartesian(xlim = c(0, max_x*1.1), ylim = c(0,20)) # Use coord_cartesian for setting limits
  
  if(!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE) #Create output directory if doesn't exist
  }
  ggsave(filename = file.path(output_dir, paste0(gsub(" ", "_", species), "_from_all_ARMS_locations", ".png")), 
         plot = plot, width = 2400, height = 1200, units = "px", dpi = 300) #add plot into right directory
  return(plot)
}
```

## Function for graph of sea distances

This function takes the same arguments as the graph before but also a location needs to be provided. This graphs consist of one species that is present in a certain ARMS location and the distance to a certain GBIF location.
Goal of graph: See the distribution of one species from one ARMS location to another location in GBIF

```{r pressure, echo=FALSE}
plot.dist.sea <- function(species, location, distances, output_dir) {
  max_x <- max(sea_loc_data$x)
  # make histograms of distances per species, with filtering on distance limit 40000
  plot <- ggplot(sea_loc_data, aes(x = x, fill = location)) +
    geom_histogram(aes(y = after_stat(count / sum(count) * 100)), 
                   binwidth = 50,
                   boundary = 0,
                   position = "stack",
                   alpha = 0.7 ) +  # Transparency level
    labs(title = paste0("Distribution of ", species, " from ", loc), x = "Sea distance (km)", y = "Percentage of total occurrences (%)") +
    theme_minimal()+
    #scale_fill_brewer(palette = "Set1") +
    theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"), 
          plot.margin = margin(0.3, 0.3, 0.4, 0.4, "cm"),
          axis.text = element_text(size = 10),           
          axis.title = element_text(size = 12),
          axis.text.x = element_text(angle = 45, hjust = 1)) +         
    scale_x_continuous(breaks = seq(0, max_x*1.1, by = 500), expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    coord_cartesian(xlim = c(0, max_x*1.1), ylim = c(0,100))
  
  
  if(!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  ggsave(filename = file.path(output_dir, paste0(gsub(" ", "_", species), "_from_", location, ".png")), 
         plot = plot, width = 2000, height = 1200, units = "px", dpi = 300)
  return(plot)
}
```
## Function for seaway and geodesic graphs
This function takes the same arguments as the last graph but now it uses a different data frame where seaway and geodesic data is combined. Each distance type is colored in another color.
Goal of the graph: See the distribution of seaway and geodesic distances for one species at one ARMS location 
```{r}
# Make combined histogram of sea distances and fly distances
plot.dist.both <- function(species, location, distances, output_dir) {
  combined_distances <- rbind(sea_loc_data, geo_loc_data)
  max_x <- max(sea_loc_data$x)
  plot <- ggplot(combined_distances, aes(x = x, fill = DistanceType)) +
    geom_histogram(aes(y = after_stat(count / sum(count) * 100)),
                   binwidth = 50,
                   color="#e9ecef", #edges of the bars
                   alpha=0.6,
                   position = 'identity') + # Place all the bars exactly where they exactly fall on x-axis
    theme_bw() +
    ggtitle(paste0("Distribution of ", species," from sea and fly distances")) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), # set title font size, placement
      plot.margin = margin(0.3, 0.3, 0.4, 0.4, "cm"),
      axis.text = element_text(size = 10),  # Set font size for axis numbers
      axis.title = element_text(size = 12), # Set font size for title
      legend.title = element_text(size = 13, face="bold"), # Settings for legend title
      legend.text = element_text(size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1)) +  # settings for legend text
    scale_x_continuous(breaks = seq(0, max_x*1.1, by = 500), expand = c(0, 0)) +  # settings for x axis
    scale_y_continuous(expand = c(0, 0)) +
    # used expand to make sure the axes are on the lines of the axes and not above them floating
    coord_cartesian(xlim = c(0, max_x*1.1), ylim = c(0,100)) + # Use coord_cartesian for setting limits
    # set legend title and labels
    labs(x = "Sea distance (km)", y = "Percentage of total occurrences (%)", fill = "DistanceType")
  
  if(!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  ggsave(filename = file.path(output_dir, paste0(gsub(" ", "_", species), "_from_", location, "_seadist&geodesic.png")), 
         plot = plot, width = 2400, height = 1200, units = "px", dpi = 300)
  return(plot)
}
```
## Function for distrubtion of species coloured by country
This function is the same as the previous one but colors based on GBIF countries.
Goal of the graph: Show distribution of one species at one ARMS location to other GBIF locations (countries) colored in a seperate color for each GBIF location.
```{r}
plot.dist.by.country <- function(species, location, distances, output_dir) {
  max_x <- max(sea_loc_data$x)
  plot <- ggplot(sea_loc_data, aes(x = x, fill = country)) +
    geom_histogram(aes(y = after_stat(count / sum(count) * 100)),
                   binwidth = 50,
                   boundary = 0,
                   position = "stack") +  
    labs(title = paste0("Frequencies of Sea distances/country for ", species," in ", location),
         x = "Sea distance (km)", y = "Percentage of total occurrences (%)") +
    theme_bw() +
    #scale_fill_brewer(palette = "Set1") +  
    theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), 
          plot.margin = margin(0.3, 0.3, 0.4, 0.4, "cm"),
          axis.text = element_text(size = 10),           
          axis.title = element_text(size = 12),
          legend.title = element_text(size = 14), 
          legend.text = element_text(size = 12),
          legend.key.size = unit(1.5, "lines"),
          axis.text.x = element_text(angle = 45, hjust = 1)) +  
    scale_x_continuous(breaks = seq(0, max_x*1.1, by = 500), expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    coord_cartesian(xlim = c(0, max_x*1.1), ylim = c(0,100)) 
  
  if(!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  ggsave(filename = file.path(output_dir, paste0(gsub(" ", "_", species), "_from_", location, "_by_country.png")), 
         plot = plot, width = 2400, height = 1200, units = "px", dpi = 300)
  return(plot)
}
```
## Function for graph coloured based on occurence year
### First check if year falls into categories
```{r}
assign_year_category <- function(year) {
  if (is.na(year)) {
    return(NA)   # return NA when year is not present
  }
  for (category in year_categories) {
    range <- as.numeric(unlist(strsplit(category, "-"))) # save years as numeric without "-"
    if (year >= range[1] & year < range[2]) {  # if the year falls into this category
      return(category) # return this category
    }
  }
  return(NA) # If year doesn't fall into any category, return NA
}
```
### Plot function
Takes the same arguments as the other graphs but colors on the year of occurrence.
```{r}
plot.dist.by.year <- function(species, location, distances, output_dir) {
  plot <- ggplot(sea_loc_data, aes(x = x, fill = year_category)) +
    geom_histogram(aes(y = after_stat(count / sum(count) * 100)),binwidth = 50, boundary = 0, position = "stack") +  # adjust the binwidth to personal preference
    labs(title = paste0("Frequencies of Sea distances/year for ", species," in ", location),
         x = "Sea distance (km)", y = "Percentage of total occurrences (%)") +
    theme_bw() +
    scale_fill_brewer(palette = "YlOrRd", na.value = "black") + # You can choose a different palette if you like
    theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"), 
          plot.margin = margin(0.3, 0.3, 0.4, 0.4, "cm"),
          axis.text = element_text(size = 10),           
          axis.title = element_text(size = 12),
          legend.title = element_text(size = 10),   
          legend.text = element_text(size = 8),    
          legend.key.size = unit(1.5, "lines"),
          axis.text.x = element_text(angle = 45, hjust = 1)) +   
    scale_x_continuous(breaks = seq(0, 7500, by = 500), expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    coord_cartesian(xlim = c(0, 7500), ylim = c(0,100)) 
  
  if(!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  ggsave(filename = file.path(output_dir, paste0(gsub(" ", "_", species), "_from_", location, "_by_year.png")), 
         plot = plot, width = 2400, height = 1200, units = "px", dpi = 300)
  return(plot)
}
```
## Summary
All goals of these graphs is to see a certain pattern of how a species behave in terms of movement and where is comes from. This can than be used to give a certain label to a species.
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
